Exploring Compiler Optimization Techniques

### 2014-11-13 doc/paper.tex
Renamed phiability to SSA conversion, and added more introduction

Code optimization is most often done by first translating the source code
(as read by a \textit{lexer} and interpreted by a \textit{parser}) into an
\textit{intermediate form}, which is subsequently converted into assembly. Such
an intermediate form must be designed in such a way that it can be analyzed and
optimized easily.\cite{dragon}

@@

  \section{Phiability optimization}
  \subsection{Problem}
Imperative languages usually allow variables and memory to be rewritten. SSA 
inherently, does not allow temporary variables to change their value over time. 
SSA supports several ways to support this model. One is using the \verb+alloca+
instruction to allocate memory and use the \verb+load+/\verb+store+ system,
another, is cleverly using $\phi$ (phi) nodes.

@@

A system that would allow turning this \verb+alloca+ system into a more
SSA-appropriate  system would get rid of all these cases where alloca would form
an exception, by morphing an advanced \verb+load+/\verb+store+ mechanism into a
mechanism using mostly temporaries for storage.

@@

Can't use a $\phi$ node system, since it needs \verb+i+ to actually exist in 
memory. It is therefore hard for a front-end to decide which system to use, and 
many\footnote{At least clang does so: \textit{echo "void foo() \{ int a; \}" |
 clang -x c - -S -emit-llvm -o /dev/stdout}}
default to using the first system all of the time, relying on the 
middle-end to optimize it into a $\phi$ node system. If an \verb+alloca+ 
variable can convert its \verb+load+/\verb+store+ system it shall be considered 
\textit{phiable}.

@@

Can \verb+t1+ be considered phiable? It's been established that an \verb+alloca+ system 
is not phiable if the memory is actually required to exist. This means (naively) 
that a system is not phiable if the \verb+alloca+ instruction is used outside 
its own \verb+load+/\verb+store+ instructions: that is if it is ever used in an 
instruction, except as the first operand of a \verb+store+ or \verb+load+.

>>>>

Code optimization is most often done by first translating the source code
(as read by a \textit{lexer} and interpreted by a \textit{parser}) into an
\textit{intermediate form}, which is subsequently converted into assembly.\cite{dragon}
Such an intermediate form must be designed in such a way that it can be analyzed and
optimized easily. Such a intermediate form commonly consists of a
list of instructions (using so-called \textit{three-way address code}) grouped
into \textit{basic blocks}. Each basic block has one starting point and one
end-point. The endpoint is always a \textit{terminal instruction}
which causes control flow to leave the block (\verb+jmp+, \verb+split+, \verb+ret+
and \verb+leave+ in the intermediate form used here).

@@

  \section{SSA conversion optimization}
  \subsection{Problem}
Imperative languages usually allow variables and memory to be rewritten. SSA 
inherently, does not allow temporary variables to change their value over time. 
SSA supports several ways to support this model. One is using the \verb+alloca+
instruction to allocate memory and use the \verb+load+/\verb+store+ system,
another, is cleverly using SSA and its $\phi$ nodes. The first translation uses
rewritable memory and is therefore not a strict SSA representation of control
flow.\footnote{The intermediate form used is still SSA compliant, however, even though it
allows rewriting of \textit{memory}. The temporary values (t1 etc.) are
still single-assignment. Allowing rewritable memory is a necessary evil for
imperative languages.}

@@

For further optimization it's far more convenient to turn such a complex system
with \verb+alloca+, \verb+load+ and \verb+store+ into an pure system where each
variable is truly written to once.

@@

Can't use a $\phi$ node system, since it needs \verb+i+ to actually exist in 
memory. It is therefore hard for a front-end to decide which system to use, and 
many\footnote{At least clang does so: \textit{echo "void foo() \{ int a; \}" |
 clang -x c - -S -emit-llvm -o /dev/stdout}}
default to using the first system all of the time, relying on the 
middle-end to optimize it into an SSA system. If an \verb+alloca+ 
variable can convert its \verb+load+/\verb+store+ system it shall be considered 
\textit{SSA-capable}.

@@

Can \verb+t1+ be considered SSA-capable? It's been established that an \verb+alloca+ system 
is not SSA-capable if the memory is actually required to exist. This means (naively) 
that a system is not SSA-capable if the \verb+alloca+ instruction is used outside 
its own \verb+load+/\verb+store+ instructions: that is if it is ever used in an 
instruction, except as the first operand of a \verb+store+ or \verb+load+.
